---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# Pattern Causality

<!-- badges: start -->
<!-- badges: end -->

The goal of patterncausality is to measure the causality in the complex system and financial market, 

## Installation

You can install the development version of patterncausality from [GitHub](https://github.com/) with:

```r
# install.packages("devtools")
devtools::install_github("skstavroglou/pattern-causality")
```

## Example

### Application in climate

We can import the existing data.

```{r}
library(patterncausality)
data(climate)
```

This dataset contains 4 time series of climate index, we could use the patterncausality in this dataset. 

Then we need to determine the `E` and `tao`.

```{r eval=FALSE}
dataset <- climate[, -1] # remove the date column
parameter <- optimalParametersSearch(Emax = 5, tauMax = 5, metric = "euclidean", dataset = dataset)
```

After the parameters are confirmed, we could calculate the pattern causality.

```{r}
X <- climate$AO
Y <- climate$AAO
pc <- PC.Mk.II.Lightweight(X, Y, E = 3, tau = 1, metric = "euclidean", h = 2, weighted = TRUE)
print(pc)
```

Then the percentage of each status will be showed below.

If we wonder the status in each time point, we can run the code.

```{r}
X <- climate$AO
Y <- climate$AAO
detail <- PC.Mk.II.Full.Details(X, Y, E = 2, tau = 1, metric = "euclidean", h = 3, weighted = TRUE)
predict_status <- detail$spectrumOfCausalityPredicted
real_stattus <- detail$spectrumOfCausalityReal
```

Then the status series will be saved in `predict_status` and `real_status`.

### Application in financial market

First of all, we can import the data of AAPL and MSFT.

```{r}
data(stock)
```

We can visualize this stock price.

```{r}
library(ggplot2)
df = data.frame(Date = as.Date(rownames(stock)), 
                Value = c(stock$AAPL.Close,
                          stock$MSFT.Close),
                Type = c(rep("AAPL", dim(stock)[1]),
                         rep("MSFT", dim(stock)[1])))
ggplot(df) +
  geom_line(aes(Date, Value, group=Type, colour = Type), linewidth=0.4) +
  theme_bw(base_size = 12, base_family = "Times New Roman") + xlab("Time") + ylab("Stock Price")+
  theme(legend.position = c(0.1,0.85), legend.box.background = element_rect(fill=NA,color = "black",linetype = 1), legend.key = element_blank(), 
        legend.title=element_blank(), legend.background = element_blank(), axis.text   = element_text(size = rel(0.8)), 
        strip.text  = element_text(size = rel(0.8)))+
  scale_color_manual(values=c("#DC143C","#191970"))
```

Then search the best parameters for the PC.

```{r eval=FALSE}
dataset <- stock
parameter <- optimalParametersSearch(Emax = 5, tauMax = 5, metric = "euclidean", dataset = dataset)
```

After that, calculate the causality of each status.

```{r}
X <- stock$AAPL.Close
Y <- stock$MSFT.Close
pc <- PC.Mk.II.Lightweight(X, Y, E = 3, tau = 1, metric = "euclidean", h = 2, weighted = TRUE)
print(pc)
```

Lastly we can also visualize this result.

```{r}
library(ggplot2)
df = data.frame(
  name=stringr::str_to_title(c(colnames(pc))),
  val=as.vector(unlist(pc))
)

ggplot(df, aes(x=name, y=val, fill=name)) +
  geom_bar(stat="identity", alpha=.6, width=.4) +
  scale_fill_grey(start=0, end=0.8) +  # start and end define the range of grays
  labs(x='Status',y='Strength')+
  theme_bw(base_size = 12, base_family = "Times New Roman") +
  theme(legend.position="none", axis.text   = element_text(size = rel(0.8)), 
                strip.text  = element_text(size = rel(0.8)))
```



## References

Based on the following **references**.

- Stavroglou, S. K., Pantelous, A. A., Stanley, H. E., & Zuev, K. M. (2019). Hidden interactions in financial markets. *Proceedings of the National Academy of Sciences, 116(22)*, 10646-10651.

- Stavroglou, S. K., Pantelous, A. A., Stanley, H. E., & Zuev, K. M. (2020). Unveiling causal interactions in complex systems. *Proceedings of the National Academy of Sciences, 117(14)*, 7599-7605.

- Stavroglou, S. K., Ayyub, B. M., Kallinterakis, V., Pantelous, A. A., & Stanley, H. E. (2021). A novel causal risk‐based decision‐making methodology: The case of coronavirus. *Risk Analysis, 41(5)*, 814-830.

- Stavroglou, S. K., Pantelous, A. A., Ayyub, B. M., Lambert, J. H., Hall, J. W., & Stanley, H. E. (2023). Prometheus Framework for Public Policy Decisions on Environmental Health Risks. *Available at SSRN 4526256*.