---
title: "Introduction to patterncausality"
author: "Stavros Stavroglou, Athanasios Pantelous, Hui Wang"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Introduction to patterncausality}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  warning = FALSE,
  collapse = TRUE,
  comment = "#>"
)
```

Pattern Causality is a novel approach for detecting and analyzing causal relationships in complex systems. This method excels at:

1. Identifying hidden patterns in time series data
2. Quantifying different types of causality (positive, negative, and dark)
3. Providing robust statistical analysis of causal relationships

The algorithm is particularly effective for:
- Financial market analysis
- Climate system interactions
- Medical diagnosis
- Complex system dynamics


## PC in financial market

First of all, we can import the data of AAPL and MSFT, also we can import data from the **yahooo** api.

```{r message = FALSE}
library(patterncausality)
data(DJS)
head(DJS)
```

We can visualize this stock price.

```{r echo=FALSE}
library(ggplot2)
df <- data.frame(
  Date = as.Date(DJS$Date),
  Value = c(
    DJS$Apple,
    DJS$Microsoft
  ),
  Type = c(
    rep("Apple", dim(DJS)[1]),
    rep("Microsoft", dim(DJS)[1])
  )
)
ggplot(df) +
  geom_line(aes(Date, Value, group = Type, colour = Type), linewidth = 0.4) +
  theme_bw(base_size = 12, base_family = "Times New Roman") +
  xlab("Time") +
  ylab("Stock Price") +
  theme(
    legend.position = c(0.25, 0.85), legend.box.background = element_rect(fill = NA, color = "black", linetype = 1), legend.key = element_blank(),
    legend.title = element_blank(), legend.background = element_blank(), axis.text = element_text(size = rel(0.8)),
    strip.text = element_text(size = rel(0.8))
  ) +
  scale_color_manual(values = c("#DC143C", "#191970"))
```

### Search the parameters

Then search the best parameters for the PC.

```{r eval=FALSE}
dataset <- DJS[,-1]
parameter <- optimalParametersSearch(Emax = 5, tauMax = 5, metric = "euclidean", dataset = dataset)
```

### Calculate the causality

After that, calculate the causality of each status.

```{r}
X <- DJS$Apple
Y <- DJS$Microsoft
pc <- pcLightweight(X, Y, E = 3, tau = 2, metric = "euclidean", h = 1, weighted = TRUE)
print(pc)
```

Lastly we can also visualize this result.

```{r}
plot_total(pc)
plot_components(pc)
```

### Full details

The details could be found with following code.

```{r eval=FALSE}
X <- DJS$Apple
Y <- DJS$Microsoft
detail <- pcFullDetails(X, Y, E = 3, tau = 2, metric = "euclidean", h = 1, weighted = TRUE)
# Access the causality components
causality_real <- detail$causality_real
causality_pred <- detail$causality_pred
print(causality_pred)
```

So far, the whole process of this algorithm has finished.

